---
title: "Effect of severe weather events in the US from 1990 to 2011"
author: "Alberto Jaimes"
date: "January 2018"
output: html_document
---

### Sypnosis  


### Data Processing  


#### Settings  

```{r setup, echo = TRUE, message = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE) 
library(R.utils)   # function to unzip data
library(tidyverse) # useful packages for data science     
library(janitor)   # clean variable names
library(lubridate) # makes easy handle dates
```


#### Data downloading  

```{r data_downloading}
url_nooa <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
path_compressed   <- "Storm_data.csv.bz2"
path_uncompressed <- "Storm_data.csv"
# download data if they are not in the working directory, and uncompress them
if (!file.exists(path_compressed)) {
  download.file(url = url_nooa, destfile = path_compressed)
  R.utils::bunzip2(path_compressed, destname = path_uncompressed, remove = FALSE)
}
```


#### Data loading  

```{r data_reading, message = FALSE}
# loading, then cleaning names
storm_raw <- readr::read_csv(path_uncompressed) %>% clean_names()
```


#### Data structure  

```{r storm_str}
glimpse(storm_raw)
```


#### Selecting useful variables  

The variables with information of the effect are **bgn_date**, the year the event started, **evtype**, the event type, **injuries**, the amount of injured people, **fatalities**, the toll caused by the weather event, **propdmg**, property damage rounded up to three significative digits, **propdmgexp**, magnitude of the property damage, **cropdmg**, crop damage up to three signifivative digits, and **cropdmgexp**, magnitude of the crop damage.

```{r select_variables}
# select useful variables, extract year of the events, and transform character variables
storm_all <- storm_raw %>% 
  select(bgn_date, evtype, injuries, fatalities, propdmg:cropdmgexp) %>% 
  mutate(year = year(mdy_hms(bgn_date)), evtype = str_to_title(evtype), 
         propdmgexp = str_to_lower(propdmgexp), cropdmgexp = str_to_lower(cropdmgexp)) %>%
  select(-bgn_date) 
```  


#### Filter observations  

The yearly numbers of events is shown in the next plot.

```{r graph_events_year, fig.align = "center"}
storm_all %>% count(year) %>% plot(type = "b", pch = 19, cex = .5, ann = FALSE)
title(xlab = "Year", ylab = "Number of events", main = "Yearly number of events")
abline(v = 1996, col = "forestgreen")
grid()
```

The earlier years have fewer events recorded, likely due to a lack of recording; I will do the analysis with events since 1996, and with values bigger than zero.

```{r filter_records}
storm_recent <- filter(storm_all, year >= 1996, 
                       propdmg > 0 | cropdmg > 0 | injuries > 0 | fatalities > 0)
```

#### Event Type Preprocessing   

The next table displays the top ten weather events in injuries.

```{r top_ten_fatalities}
storm_recent %>%
  count(evtype, wt = injuries, sort = T) %>% rename(events = n) %>% 
  head(10) %>% knitr::kable()
```

*Ttsm Wind* and *Thunderstorm Wind* correspond to the same event type; nevertheles, their records are splited in two different *evtype*. This behaviour is repeated in other cases, e. g. *Hurricane* and *Hurricane/Typhoon*. In this step I gather the several forms of each **evtype**.  

```{r process_evtype}
patterns <- c("hurrican.+" = "hurricane", "typhoon" = "hurricane", "wind " = "winds ",
              "heavy rain.+" = "heavy rain", "rains " = "rain", "hvy" = "heavy",
              "tstm" = "thunderstorm", "thundertorm" = "thunderstorm", 
              "thunderstormw" = "thunderstorm", "thunderstorms" = "thunderstorm",
              "thundeerstorm" = "thunderstorm", "thunderestorm" = "thunderstorm",
              "\\s+" = " ", "\\." = "", "(|)" = "", "^\\s+" = "", "\\s+$" = "")
# lowercase, them replace patterns, finally coerse evtype to title
storm_final <- storm_recent %>%
  mutate(evtype = str_to_lower(evtype) %>% str_replace_all(patterns) %>% str_to_title())
```


#### Preprocess to economic effect  

Reshape storm events to tidy format:
```{r reshaping_economic}
# reshaping, tidy format
storm_economic_tidy <- storm_final %>% 
  select(-injuries, -fatalities) %>% 
  gather(type_damage, cost, propdmg, cropdmg) %>% 
  gather(type_mag, magnitude, propdmgexp, cropdmgexp)
```  

In order to obtain the actual economic effect of the weather events we have to multiply the cost by  $10 ^ {magnitude}$, the damage is displayed in 2011 dollars. First, we coerse the values in *magnitude* to the appropiate exponent: $0 = 0$, $k = 3$, $m = 6$, $b = 9$.

```{r values_magnitude}
storm_economic_tidy$magnitude %>% unique
# replace NA with zeros, coerse variables to exponents
storm_economic_middle <- storm_economic_tidy %>% 
  mutate(magnitude = coalesce(magnitude, "0"),
         magnitude = str_replace_all(magnitude, c("k" = "3", "m" = "6", b = "9")),
         magnitude = as.integer(magnitude))
``` 
Now we calculate the actual economic damage by event.
```{r calculate_economic_damage}
# function to calculate actual damage
actual_damage <- function(cost, mag) {
  if (near(cost, 0)) 0 else cost * 10 ^ mag
}
# actual damage by event
storm_economic_final <- storm_economic_middle %>% 
  mutate(damage = map2_dbl(cost, magnitude, .f = actual_damage))
```  


#### Aggregation of effect on health  

Aggregate total injuries, and fatalities by event type.

```{r aggregation_health}
storm_health <- storm_final %>% 
  group_by(evtype) %>% 
  summarise(injuries = sum(injuries, na.rm = TRUE), fatalities = sum(fatalities, na.rm = TRUE))
```


#### Aggregation of economic effect  

Total economic damage by event type and kind of damage, either property or crop.

```{r aggregation_economic}
storm_economic <- storm_economic_final %>% 
  group_by(evtype, type_damage) %>% summarise(damage = sum(damage, na.rm = TRUE)) %>% 
  arrange(desc(damage)) %>% ungroup()
```


### Results  


#### Effect on health  

From the National Weather [Service Storm data documentation][documentation] we get that the effects on health are given in the variables *fatalities*, and *injuries*, for fatalities and injuries, respectively.  

```{r effect_health}
# fatalities
storm_health %>% arrange(desc(injuries)) %>% head(10)  
# injuries
storm_health %>% arrange(desc(fatalities)) %>% head(10)
```


#### Economic Effects  


### Across the U.S., which types of events are most harmful with respect to population health?  


### Across the U.S,  which types of events have the greatest economic consequences?  


### Conclusion  


### Session info  

```{r session_info}
devtools::session_info()
```  


[documentation]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
