---
title: "Effect of severe weather events in the US from 1990 to 2011"
author: "Alberto Jaimes"
date: "January 2018"
output: html_document
---

```{r setup, include = FALSE, messsage = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(R.utils)   # function to unzip data
library(tidyverse) # useful packages for data science
library(janitor)   # clean variable names
library(lubridate) # makes easy handle dates
```

### Introduction  

### Sypnosis  

### Data Processing  

#### Data downloading  

First download the Storm data in the current directory, and uncompress them, if you haven't.

```{r data_gathering}
url_nooa <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
path_compressed   <- "Storm_data.csv.bz2"
path_uncompressed <- "Storm_data.csv"
# download data if they are not in the working directory, and uncompress them
if (!file.exists(path_compressed)) {
  download.file(url = url_nooa, destfile = path_compressed)
  R.utils::bunzip2(path_compressed, destname = path_uncompressed, remove = FALSE)
}
```

#### Data loading  

Load the data and clean names.

```{r data_reading}
storm_raw <- readr::read_csv(path_uncompressed) %>% clean_names()
```

#### Selecting useful variables  

```{r cleansing}
# select useful variables, extract year of the events, and coerse variables
storm_all <- storm_raw %>% 
  select(bgn_date, evtype, injuries, fatalities, propdmg:cropdmgexp) %>% 
  mutate(year = year(mdy_hms(bgn_date)), evtype = str_to_title(evtype), 
         propdmgexp = str_to_lower(propdmgexp), cropdmgexp = str_to_lower(cropdmgexp)) %>%
  select(-bgn_date)
```  

#### Filtering years  

The next plot shows the yearly numbers of events.

```{r graph_events_year}
storm %>% count(date) %>% plot(type = "b", pch = 19, cex = .5, ann = FALSE)
title(xlab = "Year", ylab = "Number of events", main = "Yearly number of events")
grid()
```

The earlier years have fewer events recorded, likely due to a lack of meters; I will do the analysis with events since 1996.

```{r filter_years}
storm <- filter(storm_all, year >= 1995)
```


### Results  

#### Effect on health  

From the National Weather [Service Storm data documentation][documentation] we get that the effects on health are given in the variables *fatalities*, and *injuries*, for fatalities and injuries, respectively.  

```{r effect_health}
# fatalities
storm %>% count(evtype, wt = fatalities, sort = T) %>% filter(n > 0)
# injuries
storm %>% count(evtype, wt = injuries,   sort = T) %>% filter(n > 0)
```

#### Economic Effects  

In the [documentation][documentation] we also find that the economic effects are displayed in the variables *propdmg* for the property damage, and *cropdmg* for the crop damage, in order to obtain the true damage we have to multiply they for the magnitude indicated in the *propdmg* for the property damage, and *cropdmgexp* for the crop damage, respectively, the damage is displayed in 2011 dollars.  

The conversion from the magnitued to exponents is the next:

```{r tabla_coersion}
symbol_ <- c("-", "?", "+", "h", "k", "m", "b")
value_  <- c(0, 0, 0, 100, 1000, 1e6, 1e9)
df_coersion <- tibble(symbol = symbol_, value = value_)
```

```{r actual_value}
storm %>% 
  left_join(df_coersion, by = c("propdmgexp" = "symbol")) %>%
  left_join(df_coersion, by = c("cropdmgexp" = "symbol"))
  mutate(value = coalesce(value, 0))

```


### Across the U.S., which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?  

### Across the U.S, which types of events have the greatest economic consequences?

### Conclusion  

### Session info  

```{r session_info}
devtools::session_info()
```



[documentation]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
