---
title: "Effect of severe weather events in the US from 1990 to 2011"
author: "Alberto Jaimes"
date: "January 2018"
output: html_document
---

### Introduction  

### Sypnosis  

### Data Processing  

#### Settings  

```{r setup, include = TRUE, messsage = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
library(R.utils)   # function to unzip data
library(tidyverse) # useful packages for data science
library(janitor)   # clean variable names
library(lubridate) # makes easy handle dates
```

#### Data downloading  

```{r data_downloading}
url_nooa <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
path_compressed   <- "Storm_data.csv.bz2"
path_uncompressed <- "Storm_data.csv"
# download data if they are not in the working directory, and uncompress them
if (!file.exists(path_compressed)) {
  download.file(url = url_nooa, destfile = path_compressed)
  R.utils::bunzip2(path_compressed, destname = path_uncompressed, remove = FALSE)
}
```

#### Data loading  

```{r data_reading, message = FALSE}
# loading, then cleaning names
storm_raw <- readr::read_csv(path_uncompressed) %>% clean_names()
```

#### Selecting useful variables  

The variables with information of the effect are *bgn_date*, the year the event started, *evtype*, the event type, *injuries*, the amount of injured people, *fatalities*, the amount of death people caused by the weather event, *propdmg*, property damage rounded up to three significative digits, *propdmgexp*, magnitude of the property damage, *cropdmg*, crop damage up to three signifivative digits, and *cropdmgexp*, magnitude of the crop damage.

```{r cleansing}
# select useful variables, extract year of the events, and coerse variables
storm_all <- storm_raw %>% 
  select(bgn_date, evtype, injuries, fatalities, propdmg:cropdmgexp) %>% 
  mutate(year = year(mdy_hms(bgn_date)), evtype = str_to_title(evtype), 
         propdmgexp = str_to_lower(propdmgexp), cropdmgexp = str_to_lower(cropdmgexp)) %>%
  select(-bgn_date) 
```  

#### Filtering years  

The next plot shows the yearly numbers of events.

```{r graph_events_year}
storm_all %>% count(year) %>% plot(type = "b", pch = 19, cex = .5, ann = FALSE)
title(xlab = "Year", ylab = "Number of events", main = "Yearly number of events")
abline(v = 1996, col = "forestgreen")
grid()
```

The earlier years have fewer events recorded, likely due to a lack of recording; I will do the analysis with events since 1996.

```{r filter_years}
storm_recent <- filter(storm_all, year >= 1995)
```

#### Join Event type  

Some events are splited in several ways, e. g. *Hurricane* and *Hurricane/Typhoon*, in this steps I gather many description of events in only one.  

First I lowercase all descriptions, then I remove typos, and finally remove extra espaces

```{r process_gather_events}
patterns <- c("hurrican.+" = "hurricane", "typhoon" = "hurricane", "wind " = "winds ",
              "heavy rain.+" = "heavy rain", "rains " = "rain", "hvy" = "heavy",
              "tstm" = "thunderstorm", "thundertorm" = "thunderstorm", 
              "thunderstormw" = "thunderstorm", "thunderstorms" = "thunderstorm",
              "thundeerstorm" = "thunderstorm", "thunderestorm" = "thunderstorm",
              "\\s+" = " ", "\\." = "", "(|)" = "")
# replacing
storm_final <- storm_recent %>%
  mutate(evtype = str_to_lower(evtype) %>% str_replace_all(patterns) %>% str_trim(side = "both"),
         evtype = str_to_title(evtype))
```
Aggregate total injuries, and fatalities by event type.

```{r aggregation_healt}
storm_health <- storm_final %>% 
  group_by(evtype) %>% 
  summarise(injuries = sum(injuries, na.rm = TRUE), fatalities = sum(fatalities, na.rm = TRUE)) %>% 
  filter(injuries > 0 | fatalities > 0)
```



#### Preprocess to economic effect  

For computing the economic effect some reshaping is useful:  
```{r}
storm_economic <- storm_recent %>% 
  gather(type_damage, cost, propdmg, cropdmg) %>%
  gather(type_exp, magnitude, propdmgexp, cropdmgexp)
```


### Results  

#### Effect on health  

From the National Weather [Service Storm data documentation][documentation] we get that the effects on health are given in the variables *fatalities*, and *injuries*, for fatalities and injuries, respectively.  

```{r effect_health}
# fatalities
storm_health %>% arrange(desc(injuries)) %>% head(10)
# injuries
storm_health %>% arrange(desc(fatalities)) %>% head(10)
```

#### Economic Effects  

In the [documentation][documentation] we also find that the economic effects are displayed in the variables *propdmg* for the property damage, and *cropdmg* for the crop damage, in order to obtain the true damage we have to multiply they for the magnitude indicated in the *propdmg* for the property damage, and *cropdmgexp* for the crop damage, respectively, the damage is displayed in 2011 dollars.  

The conversion from the magnitued to exponents is the next:

```{r tabla_coersion}
magnitude_ <- c("-", "?", "+", "h", "k", "m", "b")
value_  <- c(0, 0, 0, 2, 3, 6, 9)
df_coersion <- tibble(magnitude = magnitude_, value = value_)
```

```{r actual_value}
# coerse magnitude to integers
effect_economic_start <- storm_economic %>% 
  left_join(df_coersion) %>% 
  mutate(magnitude = ifelse(is.na(value), magnitude, value),
         magnitude = coalesce(magnitude, "0") %>% as.integer()) %>% 
  select(-value)
# function to compute actual cost
actual_cost <- function(mag, x) {
  if (near(x, 0)) 0 
  else x * 10 ^ mag
  }
# calculate actual cost
effect_economic_middle <- effect_economic_start %>% 
  mutate(value = map2_dbl(magnitude, cost, actual_cost))
```


### Across the U.S., which types of events are most harmful with respect to population health?  

### Across the U.S,  which types of events have the greatest economic consequences?

### Conclusion  

### Session info  

```{r session_info}
devtools::session_info()
```



[documentation]: https://d396qusza40orc.cloudfront.net/repdata%2Fpeer2_doc%2Fpd01016005curr.pdf
