---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
--- 

### Settings

```{r confi, message = FALSE}
# load used packages
library(magrittr)
library(tidyverse)
library(knitr)     
library(lubridate) 
# shown code, and cache results the default
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

#### Loading the data


```{r load_data, message = FALSE}
# readr::read_csv can read zip files     
steps_raw <- read_csv("activity.zip") %>% as.tibble()
```

### What is mean total number of steps taken per day?

The data contains missing values, I will ignore them for the moment.

```{r steps_agregation} 
# total steps per day
steps_date <- steps_raw %>% group_by(date) %>% summarise(steps = sum(steps, na.rm = TRUE))
```

The next histogram shows the total number of steps taken each day.

```{r viz_steps_day_histogram}
ggplot(steps_date, aes(steps)) + geom_histogram(binwidth = 500, fill = "firebrick") + 
  labs(x = "Steps", y = "Days", title = "Total number of steps taken each day")
```

#### Mean and median of the total number of steps taken each day

The mean of steps taken each day is `r  round(mean(steps_date$steps), digits = 2)`, and the median is `r  median(steps_date$steps)`


### What is the average daily activity pattern?

```{r interval_aggregation}
# compute mean of steps taken by day
steps_interval <- steps_raw %>% 
  group_by(interval) %>% summarise(steps = mean(steps, na.rm = TRUE))
```
```{r viz_steps_interval}
steps_interval %>% plot(type = "l", ann = FALSE)
title(xlab = "Interval", ylab = "Steps taken", main = "Mean of steps take by interval")
grid()
```

#### Most active interval
```{r top_interval}
top_interval <- steps_interval %>% arrange(desc(steps)) %>% head(1)
```

The interval `r top_interval[1, 1]` contains the maximum, on average, number of steps, `r top_interval[1, 2] %>% round(digits = 2)` steps.


### Imputing missing values

#### Missing values in the dataset

The only column with missing values is *steps*.

```{r explore_missing_values}
number_na <- steps_raw %>% filter(is.na(steps)) %>% nrow
number_na
```


There are `r number_na` missing values in the dataset.  


#### Strategy for filling in all of the missing values

I will use the median for the 5-minute interval of the same weekday across the whole period to impute missing values.


#### Dataset with the missing data filled in

```{r imputing_na}
# imputed values
steps_median <- steps_raw %>% 
  mutate(day = wday(date, label = TRUE, abbr = FALSE)) %>% group_by(interval, day) %>%
  summarise(steps_median = median(steps, na.rm = TRUE)) %>% ungroup()
# replace missing values
steps_imputed <- steps_raw %>% mutate(day = wday(date, label = TRUE, abbr = FALSE)) %>% 
  left_join(steps_median) %>% mutate(steps = ifelse(is.na(steps), steps_median, steps)) %>% 
  select(-day, -steps_median)
```

#### Histogram of the total number of steps taken each day

```{r steps_day_imputed_histogram}
# aggregate steps by day
steps_imputed_day <- steps_imputed %>%   
  count(date, wt = steps)
# viz
steps_imputed_day %>%
  ggplot(aes(n)) + geom_histogram(binwidth = 500, fill = "forestgreen") +
  labs(x = "Steps", y = "Days", title = "Total number of steps taken each day",
       subtitle = "Imputed data")
```

#### Mean and median of steps taken per day of the imputed data

The mean of the steps taken per day is `r mean(steps_imputed_day$n) %>% round(digits = 2)`, 
and the median is `r median(steps_imputed_day$n) %>% round(digits = 2)`. The median of the
imputed data is the same as the original, but the mean increased from 
`r  mean(steps_date$steps) %>% round(digits = 2)` to `r mean(steps_imputed_day$n) %>% round(digits = 2)`.  


### Differences in activity patterns between weekdays and weekends

#### New variable in the dataset with two levels – “weekday” and “weekend”

```{r is_weekend}
# add variable pointing if the day is in the weekend
steps_imputed_enriched <- steps_imputed %>% 
  mutate(day     = wday(date, label = TRUE, abbr = FALSE)) %>% 
  mutate(weekend = ifelse(day %in% c("sábado", "domingo"), "weekend", "weekday")) %>%   
  select(-day)
```

#### Average of steps taken per 5-minutes interval

```{r}
# average number of steps per interval, and weekdays and weekend
steps_imputed_weekday <- steps_imputed_enriched %>%  
  group_by(interval, weekend) %>% 
  summarise(steps = mean(steps, na.rm = TRUE) %>% ceiling())
```
```{r}
# viz  
steps_imputed_weekday %>% 
  ggplot(aes(interval, steps, color = weekend)) +
  geom_line() + facet_grid(weekend ~ .) +
  theme(legend.position = "none") +
  labs(x = "Interval", y = "Steps", title = "Mean of steps taken by interval",
       subtitle = "Imputed data")
```

